---
title: "ST 558 HW 2: Programming in Base R"
author: "Chris Aguilar"
format: html
editor: visual
---

## Task 1: Basic Vector Practice

We first create some data to work with, then we assign subject names to each vector of data.

```{r}

# Make data
pre_treatment <- c(130, 128, 116, 124, 133, 134, 118, 126, 114, 127, 141, 138, 128, 140, 137, 131, 120, 128, 139, 135)

post_treatment <- c(114, 98, 113, 99, 107, 116, 113, 111, 119, 117, 101, 119, 130, 122, 106, 106, 124, 102, 117, 113)

# create names vector
subj_names <- paste("Subject", 1:20, sep = "_")

# assign names to each set of subjects
names(pre_treatment) <- subj_names
names(post_treatment) <- subj_names

```

Next we take the difference between blood pressures `pre_treatment` and `post_treatment`, and find the average difference.

```{r}

treatment_diffs <- pre_treatment - post_treatment

avg_diff <- mean(treatment_diffs)

avg_diff
```

We now find the patients that experienced a decrease in blood pressure.

```{r}

# Find indices
bp_decrease_idx <- which(treatment_diffs > 0)

pos_change_diffs <- treatment_diffs[bp_decrease_idx]

pos_change_diffs
```

Now, we find the average positive change.

```{r}

mean(pos_change_diffs)
```

## Task 2: Basic Data Frame Practice

Creating a data frame to reflect the data above.

```{r}

blood_pressures <- data.frame(
  patient = subj_names,
  pre_bp = pre_treatment,
  post_bp = post_treatment,
  diff_bp = pre_treatment - post_treatment)

```

We'll now look at a subset of the data where `diff_bp` is negative.

```{r}

subset(blood_pressures, diff_bp < 0)
```

We now add a logical column to see which post_treatment blood pressures are less than 120, then we print the dataframe with `knittr::kable()`

```{r}

blood_pressures$post_bp_lt_120 <- blood_pressures$post_bp < 120

knitr::kable(blood_pressures)
```

## Task 3: List Practice

We now look at data where the treatment is a placebo. We then create a dataframe similar to the treatment data above, using placebo data.

```{r}

pre_placebo <- c(138, 135, 147, 117, 152, 134, 114, 121, 131, 130)

post_placebo <- c(105, 136, 123, 130, 134, 143, 135, 139, 120, 124)

placebo_bp <- data.frame(
  patient = subj_names,
  pre_bp = pre_placebo,
  post_bp = post_placebo,
  diff_bp = pre_placebo - post_placebo,
  post_bp_lt_120 = post_placebo > 120
)
```

We now place these two dataframes in a list and practice accessing list elements.

```{r}

treatments_list <- list(treatment = blood_pressures, placebo = placebo_bp)

# accessing first element, the treatment df in 3 diff ways
treatments_list[["treatment"]]
treatments_list[[1]]
treatments_list$treatment
```

Using one line of code, we access the `placebo_bp`'s `pre_bp` column.

```{r}

treatments_list$placebo$pre_bp
```

## Task 4: Control Flow Practice

We categorize post_treatment (or placebo) blood pressures and append this categorization to the data frames with a for-loop.

```{r}

# Adding status column
treatments_list$treatment$status <- character(20)
treatments_list$placebo$status <- character(20)

# Post treatment BPs and categories
for(i in 1:nrow(treatments_list$treatment)) {
  treatments_list$treatment$status[[i]] <- ifelse(
    treatments_list$treatment$post_bp[[i]] <= 120, "Optimal", 
      ifelse(
        treatments_list$treatment$post_bp[[i]] > 120 &
        treatments_list$treatment$post_bp[[i]] <= 130, "Borderline", "High"
        )
    )
}

treatments_list$treatment

# Placebo BPs and categories
for(i in 1:nrow(treatments_list$placebo)) {
  treatments_list$placebo$status[[i]] <- ifelse(
    treatments_list$placebo$post_bp[[i]] <= 120, "Optimal", 
      ifelse(
        treatments_list$placebo$post_bp[[i]] > 120 &
        treatments_list$placebo$post_bp[[i]] <= 130, "Borderline", "High"
        )
    )
}

treatments_list$placebo
```

## Task 5: Function Writing

We expect to receive similar data in the future and it can be time-consuming to manually code up solutions for each set of data. We'll be interested in different statistics for the `pre`, `post`, and `diff` columns from both data frames, so we should write a function to extract these quickly.

Our function will expect a list of dataframes with the columns `pre_bp`, `post_bp`, and `diff_bp` in each data frame. It will also take a `stat` argument, set to `"mean"` by default.

```{r}

stat_maker <- function(x, stat = "mean") {
  
  my_fun <- get(stat)
  
  # variables of interest
  vars <- c("pre_bp", "post_bp", "diff_bp")
  
  # creating output containers, one per dataframe containing one slot per variable
  treatment_stats <- vector("numeric", 3)
  placebo_stats <- vector("numeric", 3)
  
  # looping through the treatment data frame for the stat of interest
  for(i in seq_along(vars)) {
    treatment_stats[[i]] <- my_fun(x$treatment[[vars[i]]])
  }
  
  # looping through the placebo data frame for the stat of interest
  for(i in seq_along(vars)) {
    placebo_stats[[i]] <- my_fun(x$placebo[[vars[i]]])
  }
  
  # generating stat names for treatment, placebo variables of interest
  treatment_names <- paste("treatment", vars, stat, sep = "_")
  placebo_names <- paste("placebo", vars, stat, sep = "_")
  
  # combining treatment and placebo stats in order
  all_stats <- c(treatment_stats, placebo_stats)
  
  # assigning element names
  names(all_stats) <- c(treatment_names, placebo_names)
  
  all_stats
}
```

Now we use our function to compute the mean, variance, standard deviation, minimum, and maximum of each variable of interest.

```{r}

# mean, left as default
stat_maker(treatments_list)

# variance
stat_maker(treatments_list, "var")

# standard deviation
stat_maker(treatments_list, "sd")

# min
stat_maker(treatments_list, "min")

# max
stat_maker(treatments_list, "max")
```

The end. Thank you for reading! 📖
